Component response rate variation underlies the stability of complex systems

### 2020-03-11 notebook/ms.Rmd
Change the notation of minimum real part of eigenvalues

Randomly assembled complex systems can be represented as large square matrices ($\mathbf{M}$) with $S$ components (e.g., networks of species [@Allesina2012] or banks [@Haldane2011]). One element of such a matrix, $M_{ij}$, defines how component $j$ affects component $i$ in the system at a point of equilibrium [@Allesina2012]. Off-diagonal elements ($i \neq j$) therefore define interactions between components, while diagonal elements ($i = j$) define component self-regulation ($d$, e.g., carrying capacity in ecological communities). Traditionally, off-diagonal elements are assigned non-zero values with a probability $C$, which are sampled from a distribution with variance $\sigma_{M}^{2}$; diagonal elements are set to $d = -1$ [@May1972; @Allesina2012; @Allesina2015]. Local system stability is assessed using eigenanalysis, with the system being stable if the real parts of all eigenvalues ($\lambda$) of $\mathbf{M}$ are negative ($\max\left(\Re(\lambda)\right) < 0$) [@May1972; @Allesina2012]. In a large system (high $S$), eigenvalues are distributed uniformly [@Tao2010] within a circle centred at $\Re = -1$ (the mean value of diagonal elements) and $\Im = 0$, with a radius of $\sigma_{M}\sqrt{SC}$ [@May1972; @Allesina2012; @Allesina2015] (Figs 1a and 2a). Local stability of randomly assembled systems therefore becomes increasingly unlikely as $S$, $C$, and $\sigma_{M}$ increase.

@@

**Effect of $\mathbf{\sigma^{2}_{\gamma}}$ on $\mathbf{M}$ (co)variation**. The value of $\max(\Re(\lambda))$, and therefore system stability, can be estimated from five properties of $\mathbf{M}$ [@Tang2014b]. These properties include (1) system size ($S$), (2) mean self-regulation of components ($d$), (3) mean interaction strength between components ($\mu$), (4) the standard deviation between component interaction strengths ($\sigma_{M}$), and (5) the correlation of interaction strengths between components, $M_{ij}$ and $M_{ji}$ ($\rho$). Positive $\sigma_{\gamma}$ does not change $S$, nor does it necessarily change $E[d]$ or $E[\mu]$. What $\sigma_{\gamma}$ does change is the total variation in component interaction strengths ($\sigma_{M}$), and $\rho$. Introducing variation in $\gamma$ increases the total variation in the system. Variation of the off-diagonal elements in $\mathbf{M}$ is described by the joint variation of two random variables,

@@

Given $\sigma^{2}_\gamma = 0$, $\max(\Re(\lambda))$ decrease linearly with $\rho$ such that [@Tang2014c],

@@

Ten thousand numerical simulations of $\mathbf{M}$ revealed that $\sigma^{2}_{\gamma} > 0$ results in a nonlinear relationship between $\rho$ and $\max(\Re(\lambda))$, which can sometimes increase the stability of the system. Figure 2 shows a comparison of $\max(\Re(\lambda))$ across $\rho$ values when $\sigma^{2}_{\gamma} = 0$ versus $\sigma^{2}_{\gamma} = 1/3$ given $S = 25$ and $\sigma = 0.4$. For $-0.4 \leq \rho \leq 0.7$ (shaded region of Figure 2), expected $\max(\Re(\lambda))$ was lower given $\sigma^{2}_{\gamma} = 1/3$. For $\rho \geq -0.1$, the range of $\max(\Re(\lambda))$ values observed given $\sigma^{2}_{\gamma} = 1/3$ also had a lower bound, which included negative values for $\rho = -0.1$ and $\rho = 0$. Hence, across a wide range of system correlations, variation in the response rate of system components had a stabilising effect.

The stabilising effect of $\sigma^{2}_{\gamma}$ across $\rho$ increased with increasing $S$. Figure 3 shows 10000 numerical simulations of $\mathbf{M}$ for $\sigma^{2}_{\gamma} = 0$ versus $\sigma^{2}_{\gamma} = 1/3$ across increasing $S$ given $\sigma = 0.2$ ($\sigma$ has been lowered here to better illustrate the effect of $S$; note that now given $S = 25$, $1 = \sigma\sqrt{SC}$). For relatively small systems ($S \leq 25$), positive $\sigma^{2}_{\gamma}$ never decreased the expected $\max(\Re(\lambda))$. But as $S$ increased, the curvilinear relationship between $\rho$ and $\max(\Re(\lambda))$ resulted in $\mathbf{M}$ with lower expected $\max(\Re(\lambda))$ for low magnitudes of $\rho$. In turn, as $S$ increased, and systems became more complex, positive $\sigma^{2}_{\gamma}$ increased the proportion of numerical simulations that were observed to be stable (see below).

@@

If the off-diagonal elements of $\textbf{A}$ are sampled independently from an identical distribution, then  $\max(\Re(\lambda))$ for $\textbf{M} = \textbf{A}$ can be estimated from five values [@Tang2014b]. These values include (1) system size ($S$), (2) mean self-regulation of components ($d$), (3) mean interaction strength between components ($\mu$), (4) the standard deviation between component interaction strengths ($\sigma$), and (5) the correlation of interaction strengths between components, $M_{ij}$ and $M_{ji}$ ($\rho$). When investigating the effect of varying component response rate $Var(\gamma)$ on stability by defining $\textbf{M} = \gamma\textbf{A}$, $S$ remains unchanged. Further, values of $\gamma_{i}$ were sampled such that $E[d]$ and $E[\mu]$ also remained unchanged (in practice, diagonal elements of $\textbf{M}$ were standardised so that mean values were identical before and after adding $\gamma$). What $Var(\gamma)$ does change is the variation in component interaction strengths, and $\rho$.

@@

<!--- Tempting to remove this section entirely, but if there is enough space for it, then maybe retain it; reviewers did not mention a problem with it, and it's a short and relevant point to make.  In any case, it should be swapped with the section on feasibility below. --->
**Targeted manipulation of $\mathbf{\gamma}$**. To further investigate the potential of $Var(\gamma)$ to be stabilising, I used a genetic algorithm. Genetic algorithms are heuristic tools that mimic evolution by natural selection, and are useful when the space of potential solutions (in this case, possible combinations of $\gamma$ values leading to stability in a complex system) is too large to search exhaustively [@Hamblin2013]. Generations of selection on $\gamma$ value combinations to minimise $\max\left(\Re(\lambda)\right)$ demonstrated the potential for $Var(\gamma)$ to increase system stability. Across $S = \{2, 3, ..., 39, 40\}$, sets of $\gamma$ values were found that resulted in stable systems with probabilities that were up to four orders of magnitude higher than when $\gamma = 1$ (Fig. 4), meaning that stability could often be achieved by manipulating $S$ $\gamma$ values rather than $S \times S$ $\mathbf{M}$ elements (i.e., by manipulating component response rates rather than interactions between components).

@@

For each $S = \{2, 3, ..., 39, 40\}$, the genetic algorithm was run for 100000 random $\mathbf{M}$ ($\sigma = 0.4$, $C = 1$). The genetic algorithm was initialised with a population of 1000 different $\mathbf{\gamma}$ vectors with elements sampled i.i.d from $\gamma_{i} \sim \mathcal{U}(0, 2)$. Eigenanalysis was performed on the $\mathbf{M}$ resulting from each $\mathbf{\gamma}$ vector, and the 20 $\mathbf{\gamma}$ vectors resulting in $\mathbf{M}$ with the lowest $\max\left(\Re(\lambda)\right)$ each produced 50 clonal offspring with subsequent random mutation and crossover between the resulting new generation of 1000 $\mathbf{\gamma}$ vectors. Mutation of each $\gamma_{i}$ in a $\mathbf{\gamma}$ vector occurred with a probability of 0.2, resulting in a mutation effect of size $\mathcal{N}(0, 0.02)$ being added to generate the newly mutated $\gamma_{i}$ (any $\gamma_{i}$ values that mutated below zero were multiplied by $-1$, and any values that mutated above 2 were set to 2). Crossover occurred between two sets of 100 $\mathbf{\gamma}$ vectors paired in each generation; vectors were randomly sampled with replacement among but not within sets. Vector pairs selected for crossover swapped all elements between and including two $\gamma_{i}$ randomly selected with replacement (this allowed for reversal of vector element positions during crossover; e.g., $\{\gamma_{4}, \gamma_{5}, \gamma_{6}, \gamma_{7}\} \to \{\gamma_{7}, \gamma_{6}, \gamma_{5}, \gamma_{4}\}$ ). The genetic algorithm terminated if a stable $\mathbf{M}$ was found, 20 generations occurred, or if the mean $\mathbf{\gamma}$ fitness increase between generations was less than 0.01 (where fitness was defined as $W_{\gamma} = -\max\left(\Re(\lambda)\right)$ for $\mathbf{M}$).

>>>>

Randomly assembled complex systems can be represented as large square matrices ($\mathbf{M}$) with $S$ components (e.g., networks of species [@Allesina2012] or banks [@Haldane2011]). One element of such a matrix, $M_{ij}$, defines how component $j$ affects component $i$ in the system at a point of equilibrium [@Allesina2012]. Off-diagonal elements ($i \neq j$) therefore define interactions between components, while diagonal elements ($i = j$) define component self-regulation ($d$, e.g., carrying capacity in ecological communities). Traditionally, off-diagonal elements are assigned non-zero values with a probability $C$, which are sampled from a distribution with variance $\sigma_{M}^{2}$; diagonal elements are set to $d = -1$ [@May1972; @Allesina2012; @Allesina2015]. Local system stability is assessed using eigenanalysis, with the system being stable if the real parts of all eigenvalues ($\lambda$) of $\mathbf{M}$ are negative ($\Re(\lambda_{max}) < 0$) [@May1972; @Allesina2012]. In a large system (high $S$), eigenvalues are distributed uniformly [@Tao2010] within a circle centred at $\Re = -1$ (the mean value of diagonal elements) and $\Im = 0$, with a radius of $\sigma_{M}\sqrt{SC}$ [@May1972; @Allesina2012; @Allesina2015] (Figs 1a and 2a). Local stability of randomly assembled systems therefore becomes increasingly unlikely as $S$, $C$, and $\sigma_{M}$ increase.

@@

**Effect of $\mathbf{\sigma^{2}_{\gamma}}$ on $\mathbf{M}$ (co)variation**. The value of $\Re(\lambda_{max})$, and therefore system stability, can be estimated from five properties of $\mathbf{M}$ [@Tang2014b]. These properties include (1) system size ($S$), (2) mean self-regulation of components ($d$), (3) mean interaction strength between components ($\mu$), (4) the standard deviation between component interaction strengths ($\sigma_{M}$), and (5) the correlation of interaction strengths between components, $M_{ij}$ and $M_{ji}$ ($\rho$). Positive $\sigma_{\gamma}$ does not change $S$, nor does it necessarily change $E[d]$ or $E[\mu]$. What $\sigma_{\gamma}$ does change is the total variation in component interaction strengths ($\sigma_{M}$), and $\rho$. Introducing variation in $\gamma$ increases the total variation in the system. Variation of the off-diagonal elements in $\mathbf{M}$ is described by the joint variation of two random variables,

@@

Given $\sigma^{2}_\gamma = 0$, $\Re(\lambda_{max})$ decrease linearly with $\rho$ such that [@Tang2014c],

@@

Ten thousand numerical simulations of $\mathbf{M}$ revealed that $\sigma^{2}_{\gamma} > 0$ results in a nonlinear relationship between $\rho$ and $\Re(\lambda_{max})$, which can sometimes increase the stability of the system. Figure 2 shows a comparison of $\Re(\lambda_{max})$ across $\rho$ values when $\sigma^{2}_{\gamma} = 0$ versus $\sigma^{2}_{\gamma} = 1/3$ given $S = 25$ and $\sigma = 0.4$. For $-0.4 \leq \rho \leq 0.7$ (shaded region of Figure 2), expected $\Re(\lambda_{max})$ was lower given $\sigma^{2}_{\gamma} = 1/3$. For $\rho \geq -0.1$, the range of $\Re(\lambda_{max})$ values observed given $\sigma^{2}_{\gamma} = 1/3$ also had a lower bound, which included negative values for $\rho = -0.1$ and $\rho = 0$. Hence, across a wide range of system correlations, variation in the response rate of system components had a stabilising effect.

The stabilising effect of $\sigma^{2}_{\gamma}$ across $\rho$ increased with increasing $S$. Figure 3 shows 10000 numerical simulations of $\mathbf{M}$ for $\sigma^{2}_{\gamma} = 0$ versus $\sigma^{2}_{\gamma} = 1/3$ across increasing $S$ given $\sigma = 0.2$ ($\sigma$ has been lowered here to better illustrate the effect of $S$; note that now given $S = 25$, $1 = \sigma\sqrt{SC}$). For relatively small systems ($S \leq 25$), positive $\sigma^{2}_{\gamma}$ never decreased the expected $\Re(\lambda_{max})$. But as $S$ increased, the curvilinear relationship between $\rho$ and $\Re(\lambda_{max})$ resulted in $\mathbf{M}$ with lower expected $\Re(\lambda_{max})$ for low magnitudes of $\rho$. In turn, as $S$ increased, and systems became more complex, positive $\sigma^{2}_{\gamma}$ increased the proportion of numerical simulations that were observed to be stable (see below).

@@

If the off-diagonal elements of $\textbf{A}$ are sampled independently from an identical distribution, then  $\Re(\lambda_{max})$ for $\textbf{M} = \textbf{A}$ can be estimated from five values [@Tang2014b]. These values include (1) system size ($S$), (2) mean self-regulation of components ($d$), (3) mean interaction strength between components ($\mu$), (4) the standard deviation between component interaction strengths ($\sigma$), and (5) the correlation of interaction strengths between components, $M_{ij}$ and $M_{ji}$ ($\rho$). When investigating the effect of varying component response rate $Var(\gamma)$ on stability by defining $\textbf{M} = \gamma\textbf{A}$, $S$ remains unchanged. Further, values of $\gamma_{i}$ were sampled such that $E[d]$ and $E[\mu]$ also remained unchanged (in practice, diagonal elements of $\textbf{M}$ were standardised so that mean values were identical before and after adding $\gamma$). What $Var(\gamma)$ does change is the variation in component interaction strengths, and $\rho$.

@@

<!--- Tempting to remove this section entirely, but if there is enough space for it, then maybe retain it; reviewers did not mention a problem with it, and it's a short and relevant point to make.  In any case, it should be swapped with the section on feasibility below. --->
**Targeted manipulation of $\mathbf{\gamma}$**. To further investigate the potential of $Var(\gamma)$ to be stabilising, I used a genetic algorithm. Genetic algorithms are heuristic tools that mimic evolution by natural selection, and are useful when the space of potential solutions (in this case, possible combinations of $\gamma$ values leading to stability in a complex system) is too large to search exhaustively [@Hamblin2013]. Generations of selection on $\gamma$ value combinations to minimise $\Re(\lambda_{max})$ demonstrated the potential for $Var(\gamma)$ to increase system stability. Across $S = \{2, 3, ..., 39, 40\}$, sets of $\gamma$ values were found that resulted in stable systems with probabilities that were up to four orders of magnitude higher than when $\gamma = 1$ (Fig. 4), meaning that stability could often be achieved by manipulating $S$ $\gamma$ values rather than $S \times S$ $\mathbf{M}$ elements (i.e., by manipulating component response rates rather than interactions between components).

@@

For each $S = \{2, 3, ..., 39, 40\}$, the genetic algorithm was run for 100000 random $\mathbf{M}$ ($\sigma = 0.4$, $C = 1$). The genetic algorithm was initialised with a population of 1000 different $\mathbf{\gamma}$ vectors with elements sampled i.i.d from $\gamma_{i} \sim \mathcal{U}(0, 2)$. Eigenanalysis was performed on the $\mathbf{M}$ resulting from each $\mathbf{\gamma}$ vector, and the 20 $\mathbf{\gamma}$ vectors resulting in $\mathbf{M}$ with the lowest $\Re(\lambda_{max})$ each produced 50 clonal offspring with subsequent random mutation and crossover between the resulting new generation of 1000 $\mathbf{\gamma}$ vectors. Mutation of each $\gamma_{i}$ in a $\mathbf{\gamma}$ vector occurred with a probability of 0.2, resulting in a mutation effect of size $\mathcal{N}(0, 0.02)$ being added to generate the newly mutated $\gamma_{i}$ (any $\gamma_{i}$ values that mutated below zero were multiplied by $-1$, and any values that mutated above 2 were set to 2). Crossover occurred between two sets of 100 $\mathbf{\gamma}$ vectors paired in each generation; vectors were randomly sampled with replacement among but not within sets. Vector pairs selected for crossover swapped all elements between and including two $\gamma_{i}$ randomly selected with replacement (this allowed for reversal of vector element positions during crossover; e.g., $\{\gamma_{4}, \gamma_{5}, \gamma_{6}, \gamma_{7}\} \to \{\gamma_{7}, \gamma_{6}, \gamma_{5}, \gamma_{4}\}$ ). The genetic algorithm terminated if a stable $\mathbf{M}$ was found, 20 generations occurred, or if the mean $\mathbf{\gamma}$ fitness increase between generations was less than 0.01 (where fitness was defined as $W_{\gamma} = -\Re(\lambda_{max})$ for $\mathbf{M}$).
